Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-17 paper/ms.tex
a few edits from my first pass

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987,Bond:1999}, correlated instrumental noise
\citep{Gibson:2012}, spectroscopic calibration \citep{Czekala:2017,Evans:2015}
and residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released efficient and well-tested
implementations of this method written in \project{C++}, \project{Python}, and \project{Julia}.
These implementations are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

The application of GP models is generally limited to small datasets because
the computational cost of computing the inverse and determinant of the matrix
$K_\bvec{\alpha}$ scales as the cube of the number of data points $N$,
$\mathcal{O}(N^3)$.
This means that for large datasets, every evaluation of the likelihood will
quickly become computationally intractable.
In this case, standard non-linear optimization or MCMC will no longer be
practical inference methods.

@@

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{celerite-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}\eqlabel{lorentz-psd}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+{\left(\frac{\omega-d_j}{c_j}\right)}^2} +
    \frac{1}{1+{\left(\frac{\omega+d_j}{c_j}\right)}^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = d_j\,{(2\,c_j)}^{-1}$, and period
$P_j = 2\,\pi\,{d_j}^{-1}$.

@@

It's worth noting that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        a_+\,\cos(d_+\,\tau) + b_+\,\sin(d_+\,\tau) +
        a_-\,\cos(d_-\,\tau) + b_-\,\sin(d_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}

@@

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
elements are given by
\begin{eqnarray}
K_{nm} = k(\tau_{nm})
\end{eqnarray}
for $k(\tau_{nm})$ given by equation \eq{kernel-simple} could be computed
efficiently by taking advantage of the structure of this covariance function
and \citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

@@

To provide some insight for this method, we will start by working through a
simple example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

@@

A special case of the \celerite\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

@@

\end{itemize}
\Figure{sho} shows a plot of the PSD for these limits and several other values
of $Q$.
This figure demonstrates that for $Q \le 1/2$, the model has no oscillatory
behavior and that for large $Q$, the shape of the PSD near the peak frequency
approaches a Lorentzian.

@@

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010,Chaplin:2011,Chaplin:2013,Stello:2013}
and asteroseismology is a
key science driver for many of the upcoming large scale photometric surveys
\citep{Campante:2016,Rauer:2014,Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical parameters
(\dfmtodo{CITE}) and more sophisticated methods are computationally expensive
and they scale poorly to the current state-of-the-art datasets \citep{Brewer:2009}.

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, we will model the star as a mixture of stochastically driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed dataset using a Gaussian Process with a PSD given by a sum of
terms given by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model in $\mathcal{O}(N)$ operations and this can be combined
with standard non-linear optimization or posterior sampling methods to infer
constraints on the parameters for the given dataset.

To demonstrate this method, we will use a very simple heuristic model based on
\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
several nuisance parameters.
The first term is used to capture the granulation ``background'' using
\eq{granulation-psd} with two free parameters $S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
like $\log g$.

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987,Bond:1999}, correlated instrumental noise
\citep{Gibson:2012}, spectroscopic calibration \citep{Czekala:2017,Evans:2015}
and residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~---~this cost is prohibitive.

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, simple, and fast.

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance, autocorrelation, or ``kernel'' function
$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the parameters
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample from the posterior probability density.

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the computational
cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
is $\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood will
quickly become computationally intractable.
In this case, standard non-linear optimization or MCMC will no longer be
practical inference methods.

@@

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{celerite-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}\eqlabel{lorentz-psd}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+{\left(\frac{\omega-d_j}{c_j}\right)}^2} +
    \frac{1}{1+{\left(\frac{\omega+d_j}{c_j}\right)}^2}
\right] \quad.
\end{eqnarray}
This is the sum of two Lorentzian or Cauchy distributions with width $c_j$
centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = d_j\,{(2\,c_j)}^{-1}$, and period
$P_j = 2\,\pi\,{d_j}^{-1}$.

@@

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        a_+\,\cos(d_+\,\tau) + b_+\,\sin(d_+\,\tau) +
        a_-\,\cos(d_-\,\tau) + b_-\,\sin(d_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to scale
inferences to large datasets for any model where the kernel function is a
general sum or product of \celerite\ terms.

@@

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
elements are given by equation \eq{kernel-simple} could be computed
efficiently by taking advantage of the structure of this covariance function
and \citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

@@

To provide some insight for this method, we will follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and this is the model that was studied by
\citet{Rybicki:1995}.
To demonstrate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

@@

A special case of the \celerite\ model of great physical interest is a
stochastically-driven damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
If $\epsilon(t)$ is white noise, the PSD of this process is given by
\citep[for example][]{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

@@

\end{itemize}
\Figure{sho} shows a plot of the PSD for these limits and several other values
of $Q$.
From this figure, it is clear that for $Q \le 1/2$, the model has no
oscillatory behavior and that for large $Q$, the shape of the PSD near the
peak frequency approaches a Lorentzian.

@@

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
many of the upcoming large scale photometric surveys \citep{Campante:2016,
Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical parameters
(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
function in the time domain scale poorly to state-of-the-art datasets
\citep{Brewer:2009}.

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, one would model the star as a mixture of stochastically-driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed dataset using a GP where the PSD is a sum of terms given by
\eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$)
\emph{conditioned on the observed time series} in $\mathcal{O}(N)$ operations.
In other words, this provides us with a computationally efficient method for
performing rigorous probabilistic inference of asteroseismic parameters in the
time domain.
We expect that this has the potential to push asteroseismic analysis to lower
signal-to-noise datasets and we hope to revisit this idea in a subsequent
paper.
